The instability of variational quantum mechanics with Gaussian approximants is (not) entirely due to the unstable least squares problem

### 2013-06-14 paper.tex

The first attempt at this will compute the dynamics of a quartic oscillator, with Hamiltonian 
$$H=\hbar a^{2\dagger}a².$$
It will do so by the midpoint method, discretising the differential equation
$$Du(t)=f(t)$$
to the difference equation
$$v^{n+1}=v^{n-1}+2τfⁿ,$$
where $tⁿ=nτ$, $vⁿ$ is meant to approximate $u(tⁿ)$, and $fⁿ=f(vⁿ)$.  The results will be compared for three difference choices of $v$ and $f$.

In the first choice, the quantum state at time $t$ is expanded over a truncated Fock basis $|N〉$, as 
$$|ν(c(t))〉=|N〉c(t).$$
Schrödinger's equation expands to
$$|N〉Dc(t)=|D(ν\circ c)(t)〉≈{H\over i\hbar}|ν(c(t))〉≈|N〉〈N|{H\over i\hbar}|N〉c(t),$$
whence
$$Dc(t)≈〈N|{H\over i\hbar}|N〉c(t).$$
In this method, we set
$$vⁿ=cⁿ\qquad{\rm and}\qquad fⁿ=〈N|{H\over i\hbar}|N〉c(t),$$
solving Schrödinger's equation quite conventionally.

>>>>

The first experiment computed the dynamics of a quartic oscillator, with Hamiltonian 
$$H=\hbar a^{2\dagger}a²,$$
solving Schrödinger's equation quite conventionally.  The quantum state at time $t$ was expanded over a truncated Fock basis $|N〉=\pmatrix{|1\rangle & |2\rangle & ⋯ & |N\rangle}$, with a column vector of coefficients $c(t)$, as
$$|ν(c(t))〉=|N〉c(t).$$
Schrödinger's equation expands to
$$|N〉Dc(t)=|D(ν\circ c)(t)〉={H\over i\hbar}|ν(c(t))〉=|N〉〈N|{H\over i\hbar}|N〉c(t),$$
whence
$$Dc(t)=-i〈N|{H\over \hbar}|N〉c(t).$$
Usually, these expansions would be approximate due to basis truncation.  In this case, the Fock states diagonalise $H$, so they are exact.  However, in general the initial state can only be approximately expanded over $|N\rangle$.  Time was discretised with the explicit Euler formula, which replaces the differential equation
$$Du(t)=f(u(t))$$
with a difference equation
$$v^{n+1}=vⁿ+τfⁿ,$$
where $tⁿ=nτ$, $vⁿ$ is meant to approximate $u(tⁿ)$, and $fⁿ=f(vⁿ)$.  Let
$$vⁿ=cⁿ\qquad{\rm and}\qquad fⁿ=-i〈N|{H\over\hbar}|N〉c(t).$$
Since $〈N|H|N〉$ is diagonal, the matrix product can be performed sparely as a dot product.  The results are shown below.  The Fock states are eigenstates of the Hamiltonian, so their amplitudes should be constant: in the exact solution of Schrödinger's equation, all the ratios plotted are exactly 1.

%\centerline{\XeTeXpicfile dry.pdf width \hsize}

These results are the textbook signs of stiffness.  As the time step is reduced from values on the order of 1, the discretised solutions diverge hugely, before converging at very small time steps.  In the standard theory of discrete approximations to $Du(t)=f(u(t))$, stability is determined by the eigenvalues of the linear transform $Df$.  In Schrödinger's equation, $f$ is the Hamiltonian, a linear transform, so in our expansion, $df=f=-i〈N|H/\hbar|N〉$, and the relevant eigenvalues are $-iEⁿ/\hbar=-in(n-1)$.  The stiffness of a differential equation is often measured by the ratio of the sizes of these eigenvalues.  For the quartic oscillator, some eigenvalues are zero, so this is not a sensible measure.  However, the largest eigenvalue, which often determines the timestep, increases quadratically with the size of the Fock basis as $N(N-1)$.

The stability region of the explicit Euler formula is ?.  This means ?.
